Does Anybody Really Know What Time It Is?\\ Automating the Extraction of Date Scalars

### 2015-10-15 icde2016/intro.tex
Improved the table format
1. Increased vertical padding in each table cell
2. Added space between the table and its caption.
Cross-validation

In recent years, there has been a growth of interest in data visualization technologies for human-assisted data analysis using systems such as Polaris~\cite{Stolte:2008} and Spotfire~\cite{Ahlberg:1996}. While computers can provide high-speed and high-volume data processing, humans have both domain knowledge and the ability to process data in parallel by using our visual systems. Humans also provide the definition of what is valuable in an analysis. Accordingly, hybrid human/computer analytic systems are essential to extracting knowledge of value to humans and their societies from the large amounts of data being generated today.

@@

One of the most common data preparation tasks we see users performing is the parsing of date strings into scalar date representations, and streamlining this process is the focus of this paper. We now provide some background on the use of dates in data analytics to provide context for the problem.

@@

\begin{table}[ht]
\centering
%\begin{tabular}{ |l|l|l| }
\begin{tabular}{|p{0.4\linewidth}| p{0.4\linewidth}|}
\hline
\centering
\textbf{ICU Format} & \textbf{Example}\\ \hline
\scriptsize{EEE MMM dd HH:mm:ss zzz yyyy} & \scriptsize{Fri Apr 01 02:09:27 EDT 2011}\\ \hline
\scriptsize{[dd/MMM/yyyy:HH:mm:ss} & \scriptsize{[10/Aug/2014:09:30:40}\\ \hline
\scriptsize{dd-MMM-yy hh.mm.ss.SSSSSS a} & \scriptsize{01-OCT-13 01.09.00.000000 PM}\\ \hline
\scriptsize{MM ''yyyy} & \scriptsize{01 '2013}\\ \hline
\scriptsize{MM/dd/yyyy - HH:mm} & \scriptsize{04/09/2014 - 23:47}\\ \hline
\end{tabular}
\label{tab:dateformats}
\caption{Unusual Date Formats.}
\end{table}

@@

Our solution was to develop two machine learning algorithms for automatically deriving the format string from the user's data with over 95\% parsing accuracy. This approach allows the user to simply specify "this column is a date" and the visualization system can respond quickly and accurately enough to avoid interrupting the user's cognitive flow.

@@

Specifically, our contributions include:

\begin{itemize}
\item By analyzing an on line corpus, we provide evidence that practical date parsing requires the ability to recognize hundreds of formats;
\item We show how to extend prior work on Minimum Descriptive Length structure extraction to generate a freely available date format domain language with over 95\% accuracy;
\item We describe a second Natural Language Processing technique for generating the same date format domain language with similar accuracy. We describe how the basic algorithm is extended to support grammar variants and constraints unique to date formats. The parsing algorithm is extended to compute an overall dominant pattern over a data column;
\item We describe some limitations of this domain language that would improve its utility.
\end{itemize}

>>>>

In recent years, there has been a growth of interest in data visualization technologies for human-assisted data analysis using systems such as Polaris~\cite{Stolte:2008} and Spotfire~\cite{Ahlberg:1996}. While computers can provide high-speed and high-volume data processing, humans have both the domain knowledge and the ability to process data in parallel by using their visual systems \cite{Cleveland.McGill1984,Mackinlay:1986}. Humans also provide the definition of what is valuable in an analysis. Algorithms that rely on \emph{both} the processing capability of computer systems and human feedback, are more effective in extracting useful knowledge from the large amounts of data being generated.
\vidya{still struggling to understand how interactivity is relevant to what we're doing here...}

@@

One of the most common data preparation tasks we see users performing is the parsing of date strings into scalar date representations, and streamlining this process is the focus of this paper \vidya{do we have any data to backup this?}\richard{Yes, I think Dan has some}. We now provide some background on the use of dates in data analytics to provide context for the problem.

@@

\begin{table}[ht]
\centering
\bgroup
\def\arraystretch{1.5}
\begin{tabular}{|p{0.4\linewidth}| p{0.4\linewidth}|}
\hline
\centering
\textbf{ICU Format} & \textbf{Example}\\ \hline
\scriptsize{EEE MMM dd HH:mm:ss zzz yyyy} & \scriptsize{Fri Apr 01 02:09:27 EDT 2011}\\ \hline
\scriptsize{[dd/MMM/yyyy:HH:mm:ss} & \scriptsize{[10/Aug/2014:09:30:40}\\ \hline
\scriptsize{dd-MMM-yy hh.mm.ss.SSSSSS a} & \scriptsize{01-OCT-13 01.09.00.000000 PM}\\ \hline
\scriptsize{MM ''yyyy} & \scriptsize{01 '2013}\\ \hline
\scriptsize{MM/dd/yyyy - HH:mm} & \scriptsize{04/09/2014 - 23:47}\\ \hline
\end{tabular}
\egroup
\label{tab:dateformats}
\caption{Unusual Date Formats.}
\end{table}

@@

Our solution was to develop two machine learning algorithms for automatically deriving the format string from the user's data with over 95\% parsing accuracy. We developed two algorithms because we were not aware of any previous work on this problem and we needed to be able to cross-validate our results on a large testing corpus. Either approach allows the user to simply specify ``this column is a date" and the visualization system can respond quickly and accurately enough to avoid interrupting the user's cognitive flow.

@@

Specifically, our contributions include:
\begin{itemize}
\item By analyzing an online corpus, we provide evidence that practical date parsing requires the ability to recognize hundreds of formats;
\item We show how to extend prior work on Minimum Descriptive Length structure extraction to generate a freely available date format domain language with over 95\% accuracy;
\item We describe a second Natural Language Processing technique for generating the same date format domain language with similar accuracy. We describe how the basic algorithm is extended to support grammar variants and constraints unique to date formats. The parsing algorithm is extended to compute an overall dominant pattern over a data column;
\item We provide evidence that the development of multiple, independent parsing algorithms provides an effective means of cross-validation on large corpi;
\item We describe some limitations of this domain language that would improve its utility.
\end{itemize}
